A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
tables

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by thousands of programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

@@

R and its ancestor S are languages designed, implemented and maintained by
statisticians. Originally they were designed as glue languages, languages
that would allow to read data into vectors and call statistical routines
written in Fortran. Over three decades, the languages became widely used
across many fields of science and in industry for data analysis and data
visualization; with time additional features were added.  Modern R, as a
linguistic object of study, is fascinating. It is a vectorized, dynamically
typed, lazy functional language with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.

@@

This papers is a data-driven study of what a type system for the R language
could look like. Longer term, our intention is to propose changes to
language, but for any changes to be accepted by the user community, they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

@@

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item Manual analysis of 100 functions to validate the dynamic analysis
  results.
\end{itemize}

@@

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate that in a different environment.

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added attributes. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

@@

In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
an idea, matrix/vector polymorphism seems sensible, as mathematically
vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
(e.g., \code{as.vector(matrix(2, 2, 2)) == c(2, 2, 2, 2)}), and
\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
\code{as.matrix(c(1, 2)) == matrix(data=c(1, 2))}).  \AT{TODO fix line break
  in code at end of last sentence}

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 269K & 29.7\% \\ 
  \l & 193K & 21.3\% \\ 
  \C & 96K & 10.6\% \\ 
  \L & 83K & 9.2\% \\ 
  \sF & 60K & 6.6\% \\ 
  \sN & 47K & 5.2\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \I & 30K & 3.2\% \\ 
  \sS & 16K & 1.7\% \\ 
  \D, \l & 10K & 1.1\% \\ 
   \hline
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\end{table}

@@

\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 9566 & 1.1\% \\ 
  \D, \sN & 8801 & 1\% \\ 
  \C, \sN & 5590 & 0.6\% \\ 
  \l, \sN & 5016 & 0.6\% \\ 
  \C, \D & 2887 & 0.3\% \\ 
  \D, \sF & 1608 & 0.2\% \\ 
  \l, \sF & 1345 & 0.1\% \\ 
  \C, \l & 1283 & 0.1\% \\ 
  \C, \sF & 1168 & 0.1\% \\ 
  \C, \I & 1104 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\end{table}

@@

\begin{table}[ht]\label{tab:L0argcounts}\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 852K & 94\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\label{tab:L1top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 179K & 19.7\% \\ 
  \D & 97K & 10.7\% \\ 
  \sL & 77K & 8.6\% \\ 
  \sC & 75K & 8.3\% \\ 
  \l & 67K & 7.4\% \\ 
  \sF & 61K & 6.7\% \\ 
  \sN & 50K & 5.5\% \\ 
  \M{D} & 49K & 5.5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \df & 29K & 3.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\end{table}

@@

\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}
\end{table}

@@

\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 864K & 95.3\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L1.}
\end{table}

@@

\begin{table}[ht]
\label{tab:L2top10arg}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
 \sD & 172K & 19\% \\ 
  \sL & 77K & 8.5\% \\ 
  \D & 77K & 8.4\% \\ 
  \sC & 73K & 8\% \\ 
  \sN & 50K & 5.5\% \\ 
  \sF & 45K & 5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
  \C & 25K & 2.6\% \\ 
   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Apr  5 09:50:59 2019
\begin{table}[ht]
\label{tab:toppolyL2}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
  \sC, \sD & 1431 & 0.2\% \\ 
  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
  \I, \sD & 855 & 0.1\% \\ 
  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
  \C, \D & 672 & 0\% \\ 
   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in {\bf L2}.}
\end{table}

@@

Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
Table~\ref{tbl:argcountsL2} has the answer.

\begin{table}[ht]
\label{tab:argcountsL2}
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 851K & 93.9\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in {\bf L2}.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\label{tab:packagepolysize}
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
  oshka & 4 & 38 & quoted expression expansion \\ 
  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
  yaml & 4 & 33 & convert R to YAML \\ 
  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
   \hline
\end{tabular}
\caption{5 of the most polymorphic packages (in {\bf L0}).}
\end{table}

@@

%  magrittr & 4.16 &  \\
%  types & 3.25 & Simple type annotations for R \\
%  r6extended & 3.00 & Utilities extending barebones R6 package \\
%  utils & 2.93 & Utilities for R developers \\
%  zeallot & 2.85 & Metaprogramming package for assignment \\
%  listviewer & 2.72 & R list visualizer \\
%  proto & 2.60 &  \\
%  apercu & 2.49 & Display a short view of a complex R value (list, vector, etc.) \\
%  pipeR & 2.46 & Another pipe suite \\
%  rethinker & 2.44 & RethinkDB client for R \\

%% latex table generated in R 3.5.1 by xtable 1.8-3 package
%% Thu Mar 28 16:01:37 2019
%\begin{table}[ht]
%\label{tab:bigpolyfuns}
%\centering
%\begin{tabular}{llrl}
%  \hline
%Package & Function & Size & Description \\
%  \hline
%  magrittr & \%$>$\% & 23.00 & Tidyverse pipe function \\
%  magrittr & freduce & 17.67 & Apply a list of functions sequentially \\
%  rlang & captureArgInfo & 16.50 & Capture arg info \\
%  rlang & enquo & 16.00 & Create quosure \\
%  rlang & get\_expr & 15.00 & Extract expression \\
%  data.table & replace\_dot\_alias & 13.50 & Deals with '.' in passed expr \\
%  tibble & strip\_dim & 13.00 & Removes 'dim' attribute \\
%  data.table & copy & 11.00 & Create copy of passed value \\
%  htmlwidgets & shouldEval & 11.00 & Eval passed value depending on tag \\
%  magrittr & \%$<$$>$\% & 11.00 & Like \%$>$\% but assigns \\
%   \hline
%\end{tabular}
%\caption{10 most polymorphic functions.}
%\end{table}

@@

%
%  ls trace_export_promises > promise_packages_FINAL.txt
%
%
%
%
\section{Discussion}

@@

Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
If building an annotation for such a function, the ... argument is always ascribed the \l type, as it can indeed be inhabited by anything.

@@

The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} ... .

More generally, there is an existing literature on analyzing usage patterns
of language features.  
For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers}, ...

@@

The R programming language is an example of this.
R was developed to foster an easy and interactive programming experience, and the language is not very restrictive and highly expressive.
Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
Further, R is a very slow, and the poor performance is quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.

Types are a perfect way to both make the language easier to debug and make it run faster.
For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
If we're interested in developing an easily adoptable type system for this language, we must ensure that it captures how existing code is being used.
That is precisely what we have accomplished in this work.

Throughout this paper, we have built up an understanding of types and polymorphism in R by analyzing the usage of types across a large corpus of R code.
With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
Leveraging the \genthat~\cite{issta18} tool, we analyzed tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.
Analyzing package examples mitigates the lack of end-user R code,

\AT{Need something more, stuff commented in source.}

%Our final type system, {\bf L2}, consolidated information from 3 of R's main reflective functions (\code{typeof}, \code{class}, and \code{attributes}), as well as some extra information about certain values (e.g. the type of list elements, and the length of vectorized types) to be as precise as reasonable.
%We found that a type system such as {\bf L2} captures an overwhelming majority (>90\%) of argument-level type usage, and in it a vast majority ($\approx$80\%) of functions were monomorphic.
%Among other interesting data points, we discovered that many argument positions with a vectorized type in R are monomorphic {\it and strictly scalar}, suggesting that types or annotations for scalar-only values is useful.

%
%
%
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}

>>>>

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
%\renewcommand{\l}{\ensuremath{\small L<{\mathtt ?}>}\xspace}     % List
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
it, we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by thousands of programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our results suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

@@

R and its ancestor S are languages designed, implemented, and maintained by
statisticians. Originally they were designed as glue languages, languages
that would allow to read data into vectors and call statistical routines
written in Fortran. Over three decades, the languages became widely used
across many fields of science and in industry for data analysis and data
visualization; with time additional features were added.  Modern R, as a
linguistic object of study, is fascinating. It is a vectorized, dynamically
typed, lazy functional language with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.

@@

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to propose changes to the
language, and for any changes to be accepted by the user community they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

@@

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item A set of type annotations that captures how programmers use types
in existing R code.
\end{itemize}

@@

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an abstract syntax tree, or AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate that in a different environment.

@@

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

@@

In Figure~\ref{fig:matvec}, we see that a function which can take in a vector,
but immediately transforms it (transpose, with {\tt t}) into a matrix.  As
an idea, matrix/vector polymorphism seems sensible, as mathematically
vectors {\it are} one-dimensional matrices.  R echoes this by making conversion between the two
types easy: \code{as.vector(m)} flattens a matrix \code{m} into a vector
(e.g., \code{as.vector(matrix(1, 2))} is equivalent to \code{c(1, 1)}), and
\code{as.matrix(v)} builds a {\tt length(v) x 1} matrix (e.g.,
\code{as.matrix(c(1, 2))} is equivalent to \code{matrix(data=c(1, 2))}).

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.

\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 269K & 29.7\% \\ 
  \l & 193K & 21.3\% \\ 
  \C & 96K & 10.6\% \\ 
  \L & 83K & 9.2\% \\ 
  \sF & 60K & 6.6\% \\ 
  \sN & 47K & 5.2\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \I & 30K & 3.2\% \\ 
  \sS & 16K & 1.7\% \\ 
  \D, \l & 10K & 1.1\% \\ 
   \hline
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\label{tab:L0top10arg}
\end{table}

@@

\begin{table}[ht]\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 9566 & 1.1\% \\ 
  \D, \sN & 8801 & 1\% \\ 
  \C, \sN & 5590 & 0.6\% \\ 
  \l, \sN & 5016 & 0.6\% \\ 
  \C, \D & 2887 & 0.3\% \\ 
  \D, \sF & 1608 & 0.2\% \\ 
  \l, \sF & 1345 & 0.1\% \\ 
  \C, \l & 1283 & 0.1\% \\ 
  \C, \sF & 1168 & 0.1\% \\ 
  \C, \I & 1104 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\label{tab:L0toppoly}
\end{table}

@@

\begin{table}[ht]\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 852K & 94\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\label{tab:L0argcounts}
\end{table}

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 179K & 19.7\% \\ 
  \D & 97K & 10.7\% \\ 
  \sL & 77K & 8.6\% \\ 
  \sC & 75K & 8.3\% \\ 
  \l & 67K & 7.4\% \\ 
  \sF & 61K & 6.7\% \\ 
  \sN & 50K & 5.5\% \\ 
  \M{D} & 49K & 5.5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \df & 29K & 3.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\label{tab:L1top10arg}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}
\label{tab:toppolyL1}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 864K & 95.3\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L1.}
\label{tab:argcountsL1}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
 \sD & 172K & 19\% \\ 
  \sL & 77K & 8.5\% \\ 
  \D & 77K & 8.4\% \\ 
  \sC & 73K & 8\% \\ 
  \sN & 50K & 5.5\% \\ 
  \sF & 45K & 5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
  \C & 25K & 2.6\% \\ 
   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\label{tab:L2top10arg}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Apr  5 09:50:59 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
  \sC, \sD & 1431 & 0.2\% \\ 
  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
  \I, \sD & 855 & 0.1\% \\ 
  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
  \C, \D & 672 & 0\% \\ 
   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in {\bf L2}.}
\label{tab:toppolyL2}
\end{table}

@@

Now that we have seen the most frequently appearing argument signatures, we ask: {\it how many argument positions are polymorphic?}.
Table~\ref{tab:argcountsL2} has the answer.

\begin{table}[ht]
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 851K & 93.9\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in {\bf L2}.}
\label{tab:argcountsL2}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
  oshka & 4 & 38 & quoted expression expansion \\ 
  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
  yaml & 4 & 33 & convert R to YAML \\ 
  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
   \hline
\end{tabular}
\caption{5 of the most polymorphic packages (in {\bf L0}).}
\label{tab:packagepolysize}
\end{table}

@@

@@

%
%  
%
%
%
%
\section{Discussion}

@@

Our analysis ignores these extra arguments, instead it simply flags functions which have a {\tt ...} in their definition.
If building an annotation for such a function, the {\tt ...} argument is always ascribed the \l type, as it can indeed be inhabited by anything.

@@

The idea of measuring polymorphism via some program analysis isn't new.
For instance, some work~\cite{aakerblom2015measuring} measures the polymorphism of Python programs, but with a slightly different goal than ours: authors seek to assess the need for expressiveness in type systems that are retrofit onto dynamic languages.
Interestingly, they draw some similar conclusions to us: they find that monomorphism is far more frequent than polymorphism, and even find that typically very few functions are very highly polymorphic.

More generally, there is an existing literature on analyzing usage patterns of language features.  
For instance, the dynamic features of Smalltalk were analyzed in some work~\cite{callau2011howdevelopers},

@@

The R programming language is an example of this.
R was developed to foster an easy and interactive programming experience, and the language is not very restrictive and highly expressive.
Unfortunately, these liberties are not without a cost: debugging R code is challenging, and given how easy it is to do anything in the language, it's no surprise that it's easy to make a mistake.
Further, R is very slow, and the poor performance quickly becomes prohibitive if programmers are not careful when coding analysis for large data sets.

Types are a perfect way to both make a language easier to debug and make it run faster.
For debugging, some level of static type enforcement would give programmers more direct feedback if they are misusing values.
As for performance, type information is crucial for many optimizations performed by JITs and other language implementations.
If we're interested in developing an easily adoptable type system for R language, we must ensure that it captures how existing code is being used.
That is precisely what we have accomplished in this work.

Throughout this paper, we have built up an understanding of types and polymorphism in R by analyzing the usage of types across a large corpus of R code.
With this knowledge, we highlighted some interesting patterns of type usage in the language, and developed a number of type systems with differing levels of granularity to describe and capture the use of types by programmers in existing code.
We analyzed the tests, examples, and vignettes of over ten thousand packages on CRAN (over \PERCENTCRAN of available packages), and traced over 15 million lines of R code.

We believe that a type system akin to {\bf L2} is well suited for R, as it is expressive enough to capture current use of the language without being so specific as to make everything polymorphic.
In fact, the levels of polymorphism observed were rather low overall, with only 6\% of function arguments and 21\% of functions being polymorphic.
Importantly, {\bf L2} carries enough information to be useful to R implementations and IDEs:
language implementations would surely benefit from statically-available type information, and IDEs could use the rich type and attribute information to give programmers immediate feedback during the programming process.

%
%
%
\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new,bib/gradual}
\end{document}
